\chapter{Conclusion}%
\label{ch:conclusion}

% The Conclusion chapter is typically the last chapter you write, though most often it will be the first chapter others will read (after the abstract).
% Make sure to be concise and complete, and make it not too detailed.
% No new information is presented or derived, all information should have been presented in other chapters.

% \section{Summary \& Findings}
% Here you summarise your work (the context, questions, approach), and present the findings in context with the research questions.

% \section{Contributions}
% Highlight the contributions your thesis provides.
% These are tangible results, which can live outside the context of this thesis.
% Examples are new methods, frameworks, software.
% Make sure to provide links to Git repos or datasets if applicable.

\section{Limitations \& Threats to Validity}

The evaluation has two main limitations. First, it was conducted with non-technical stakeholders only. A planned user study with technical stakeholders could not be completed within the internship window, so the technical view was assessed structurally, against the Phase~1 technical requirements, the design principles it implements, and a frozen set of worked examples, rather than through stakeholder sessions. The non-technical findings should therefore be read as evidence for the non-technical direction of the artifact, not for both views equally. Second, all participants were drawn from a single consulting organisation, so the elicited requirements and the evaluation results reflect one environment and may not transfer to others.
% No work is perfect.
% Discuss the limitations of your study.
% A research study can only be as unbiased as the researcher and the circumstances they are working with.
% The threats to validity offers insights into the variables that were encountered and how you compensated for them.

\section{Future Work}

Three extensions follow directly from the limitations above. A comparative user study with technical stakeholders would test whether the technical view supports analyst decision-making as the non-technical view was shown to support planning decisions. Data sensitivity, the one component of business impact absent from both the scanner export and the asset inventory, could be incorporated once a source is available, completing the impact assessment. Finally, replicating the study across more than one organisation would test whether the requirements and the evaluation findings generalise beyond a single consulting context.
% This is the moment to highlight all the work you wanted to do if you had more time.
% Or any new ideas you did not have the time to explore.
% This is also where you can give an idea how to address any limitations mentioned in the previous section.

