\chapter{Introduction}\label{ch:introduction}

Cybersecurity operations in modern organisations rely on toolchains that generate large volumes of technical data. In vulnerability management, for example, scanners such as Qualys produce structured findings that include identifiers, severity scores, affected assets, exploitability indicators, and remediation status. These outputs are useful for security engineers and analysts, but they are rarely produced for non-technical stakeholders such as consultants, delivery managers, sales staff, or business decision-makers. Yet these stakeholders still influence prioritisation, budgeting, planning, and communication with clients or senior management.

In practice, the gap between technical output and stakeholder understanding is usually bridged through dashboards, reports, slide decks, and manual interpretation by experts. This process is common, but it is also fragile. It depends heavily on the expertise of the person preparing the explanation, and it is often difficult to trace how a high-level recommendation relates back to the original technical evidence. As a result, the same vulnerability data may be framed differently across teams or projects, which can lead to inconsistency in risk communication and weaker decision support. Prior work in usable security shows that interpreting security information is difficult even for technical practitioners, and that tools often create additional effort when their outputs are poorly aligned with the needs of those who must act on them \cite{werlinger2008ids, cranor2008framework}.

This problem is not only a matter of interface design. From a Software Engineering perspective, it can be understood as a problem of system design, abstraction, and stakeholder-oriented representation. Software architecture is meant to structure complexity, separate concerns, and provide different views for different audiences \cite{bass2012software, kruchten1995viewmodel, parnas1972modules}. In principle, these ideas are highly relevant to cybersecurity communication. However, in practice they are usually applied to support development, maintenance, and runtime behaviour for technical users, not to systematically translate technical cybersecurity outputs into explanations that are understandable to non-technical stakeholders.

At the same time, Explainable Artificial Intelligence (XAI) has introduced methods that aim to make model outputs more understandable, including feature attribution, surrogate models, and counterfactual explanations \cite{doshivelez2017rigorous, guidotti2018survey, barredo2020xai}. These methods have expanded the discussion around transparency and interpretability, but most of this work remains model-centric. The focus is typically on explaining a prediction made by a machine learning model, rather than on designing software systems that integrate explanations into end-to-end workflows. This limits their usefulness in settings such as cybersecurity, where information often comes from multiple tools, must remain traceable to evidence, and needs to be presented differently depending on the stakeholder and decision context \cite{lipton2018mythos, miller2019explanation}.

A related line of work comes from natural language generation and data-to-text research. This literature shows how structured data can be transformed into human-readable narratives by selecting relevant content, aggregating details, and presenting information in a form that supports interpretation and decision-making \cite{gatt2018survey, reiter2000nlg}. Although this work is not specific to cybersecurity, it is relevant because it treats explanation as a design problem rather than as a purely algorithmic one. Together, these strands suggest that explainability can be treated as a system-level concern that can be operationalised through architectural abstractions and role-specific representations.

This thesis therefore studies the design of software architectures that translate vulnerability management outputs into stakeholder-oriented explanations. The work is carried out in collaboration with Accenture in the domain of cyber defence and cyber resilience. Because of confidentiality constraints, the project does not use customer-specific or sensitive data. Instead, it focuses on a representative vulnerability management workflow based on synthetic data modelled on Qualys-style exports. The emphasis is not on improving vulnerability detection itself, but on designing abstractions, architectural responsibilities, and explanation mechanisms that make technical findings more understandable and useful for non-technical stakeholders.

The main research question of this thesis is:

\textit{How can software abstractions incorporating explainability concepts be designed within a software system to translate vulnerability scanner outputs into stakeholder-oriented explanations for non-technical audiences?}

To answer this question, the thesis addresses the following sub-questions:

\begin{itemize}
    \item \textbf{RQ1:} Which requirements and quality attributes are needed for effective stakeholder-oriented explanations of vulnerability management outputs, and how can these be elicited systematically?
    \item \textbf{RQ2:} How can an architectural design and abstraction model be constructed to transform Qualys-style vulnerability findings into explainable stakeholder-specific representations?
    \item \textbf{RQ3:} To what extent does the proposed design improve perceived understandability and usefulness compared to baseline vulnerability reporting, and what trade-offs emerge between explanation richness, correctness, and maintainability?
\end{itemize}

This research follows a Design Science Research approach, which is well suited to Software Engineering projects in which the main contribution is an artifact that is designed and evaluated in context \cite{hevner2004dsr}. The artifact in this thesis consists of a conceptual software architecture and a prototype that implement a stakeholder-oriented explanation pipeline for vulnerability management. The evaluation focuses on whether the resulting explanations are understandable, useful, and traceable to the underlying technical evidence.

The contribution of this thesis is threefold. First, it proposes a Software Engineering view on explainability in cybersecurity by treating explanation as an architectural concern rather than as an isolated interface feature or model add-on. Second, it delivers a concrete architecture and prototype for transforming technical vulnerability outputs into stakeholder-oriented representations. Third, it evaluates this design with respect to its practical usefulness for supporting understanding and communication in a vulnerability management workflow.

The remainder of this thesis is organised as follows. Chapter~\ref{ch:background} provides the conceptual background on vulnerability management, software architecture, explainability, and human-centred security. Chapter~\ref{ch:related} discusses related work and positions the thesis with respect to the current state of the art. Chapter~\ref{ch:methods} presents the methodology, followed by the design and implementation of the artifact in Chapter~\ref{ch:designandprototype}, the evaluation in Chapter~\ref{ch:analysis}, and the conclusions.